---
title: "Fonology package"
author: "Guilherme D. Garcia | [gdgarcia.ca](https://gdgarcia.ca)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fonology package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Distinctive features

### From phonemes to features

The function `getFeat()` requires a set of phonemes `ph` and a language `lg`. It outputs the minimal matrix of distinctive features for `ph` given the phonemic inventory of `lg`. Four languages are supported: English, French, Portuguese, and Spanish. Here are some examples.

```{r, message=F}
library(Fonology)

getFeat(ph = c("i", "u"), lg = "English")
getFeat(ph = c("i", "u"), lg = "French")
getFeat(ph = c("i", "y", "u"), lg = "French")
getFeat(ph = c("p", "b"), lg = "Portuguese")

```

### From features to phonemes

The function `getPhon()` requires a feature matrix `ft` and a language `lg`. It outputs the set of phonemes represented by `ft` given the phonemic inventory of `lg`. The languages supported are the same as those supported by `getFeat()`.

```{r, message=F}
library(Fonology)

getPhon(ft = c("+syl", "+hi"), lg = "French")
getPhon(ft = c("-DR", "-cont", "-son"), lg = "English")
getPhon(ft = c("-son", "+vce"), lg = "Spanish")

```


## IPA transcription for Portuguese

The function `ipa_pt()` takes a monomorphemic Portuguese `word` (real or not) in its orthographic form and returns its phonemic (i.e., broad) transcription based on **Brazilian Portuguese**. The narrow transcription can be generated adding `narrow = T` to the function. 

```{r, message=F}
library(Fonology)

ipa_pt("atletico") # existing word
ipa_pt("cantalo", narrow = T) # nonce word with nasalized vowel
ipa_pt("teatro")
ipa_pt("antidepressivo", narrow = T) # secondary stress in narrow transcription
ipa_pt("feris") # nonce word with potential penultimate stress despite final syllable
ipa_pt("tialo")
ipa_pt("piato", narrow = T) # diphthongization expected in narrow transcription

```

### A note on stress

Stress is assigned based on two scenarios. First, **real** words (non-verbs) have their stress assignment derived from the [Portuguese Stress Lexicon](https://gdgarcia.ca/psl.html) (if the word is listed there). Second, **nonce** words follow the general patterns of Portuguese stress *as well as* probabilistic tendencies shown in my work (Garcia, [2017a](https://gdgarcia.ca/thesis.html), [2017b](https://gdgarcia.ca/garcia_phon_2017.html), [2019](https://gdgarcia.ca/garcia_lg_2019.html)). As a result, a nonce word *may* have antepenultimate stress under the right conditions based on lexical statistics in the language. Likewise, words with other so-called exceptional stress patterns are also generated probabilistically (e.g., `LH]` words with penultimate stress). Stress and weight are also used to apply both spondaic and dactylic lowering to narrow transcriptions, following work such as Wetzels (2007). Secondary stress is provided when `narrow = T`.

### A note on key assumptions

There are several assumptions about surface-forms when `narrow = T`. Most of these assumptions can (and probably will) be adjusted as the package improves its accuracy and coverage. Diphthongization, for example, is sensitive to phonotactics. A word such as `CV.ˈV.CV` will be narrowly transcribed as `ˈCGV.CV` (except when the initial consonant is an affricate (allophonic), which *seems* to lower the probability of diphthongization based on my judgement). Diphthongization is not applied if the onset is complex. Needless to say, these assumptions are based on a particular dialect of Brazilian Portuguese, and I do not expect all of them to seamlessly apply to other dialects (although some assumptions are more easily generalizable than others).

Narrow transcription also includes (final) vowel reduction, voicing assimilation, l-vocalization, vowel devoicing, palatalization, and epenthesis in `sC` clusters and other consonant sequences that are expected to be repaired on surface forms (e.g., *kt*, *gn*). Examples can be generated with the function `test_pt()`.

### Helper functions

If you plan to tokenize texts and create a table with individual columns for stress and syllables, you can use some simple additional helper functions. For example, `getWeight_pt()` will take a syllabified word in Portuguese and return its weight profile (e.g., `getWeight_pt("kon.to")` will return `HL`). The function `getStress()` will return the stress position of a given word (up to preantepenultimate stress)---the word must already be stressed, but the symbol used can be specified in the function (argument `stress`). Finally, `getSyl()` will extract a given syllable from a string. For example, `getSyl(word = "kom-pu-ta-doɾ", pos = 3, syl = "-")` will take the antepenultimate syllable of the string in question. The default symbol for syllabification is the period. Finally, `dePlu_pt()` will remove plural `s` from words that are listed in the Portuguese Stress Lexicon (to avoid treating morphemic `s` as a real coda word-finally)---note that `ipa_pt()` is not supposed to be used with polymorphemic strings. 


## Plotting vowels

The function `plotVowels()` creates a vowel trapezoid using `ggplot2` and also returns the LaTeX code to create the same trapezoid using the [`vowel` package](https://ctan.org/pkg/vowel?lang=en).

## Bigram probabilities

The function `biGram_pt()`returns the log bigram probability for a possible `word` in Portuguese (`word` must be broadly transcribed). The string must use broad phonemic transcription, but no syllabification or stress. Diacritics allows: `~` for `a` and `w` in nasal diphthongs. The reference used calculate probabilities is the [Portuguese Stress Lexicon](https://gdgarcia.ca/psl.html).


## Generate hypothetical words in Portuguese

The function `gen_pt()` generates a hypothetical word in Portuguese. Note that this function is meant to be used to get you started with nonce words. You will most likely want to make adjustments based on phonotactic preferences. The function already takes care of *some* OCP effects and it also prohibits more than one onset cluster per word, since that's realtively rare in Portuguese. Still, there will certainly be other sequences that sound less natural. The function is not too strict because you may have a wide range of variables in mind as you create novel words.

```{r, message=FALSE}
library(Fonology)
# gen_pt(profile = "LHL")
# 
# tibble(word = character(5)) %>% 
#   rowwise() %>% 
#   mutate(word = gen_pt("LHL"))
```

## References

- Garcia, G. D. (2014). Portuguese Stress Lexicon. Available at [gdgarcia.ca/psl.html](https://gdgarcia.ca/psl.html).

- Garcia, G. D. (2017). Weight effects on stress: Lexicon and grammar [PhD thesis, McGill University]. https://doi.org/10.31219/osf.io/bt8hk

- Garcia, G. D. (2017). Weight gradience and stress in Portuguese. Phonology, 34(1), 41–79. https://doi.org/10.1017/S0952675717000033

- Garcia, G. D. (2019). When lexical statistics and the grammar conflict: Learning and repairing weight effects on stress. Language, 95(4), 612–641. https://doi.org/10.1353/lan.2019.0068

- Wetzels, L., (2007) “Primary Word Stress in Brazilian Portuguese and the Weight Parameter”, Journal of Portuguese Linguistics 6(1), 9-58. doi: https://doi.org/10.5334/jpl.144


